### 2017-01-20 CO750-ApproximationAlgorithmsTopics/lectures/lecture_05.tex
partial lecture 5 scribe

\lhead{2017-01-17}
\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 4} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
Recall from Lecture $3$:
\begin{theorem}
Given any metric $(V,d)$ one can randomly find a tree $T$ such that
\begin{enumerate}
\item $d(u,v) \leq d^T(u,v)$ for all $u,v\in V$
\item $E[d^T(u,v)] \leq O(\log n) d(u,v)$ for all $u,v\in V$.
\end{enumerate}
\end{theorem}
\begin{proof}
We may assume $\max_{u,v \in V} d(u,v) = 2^\delta$ for some $\delta \in \N$ and $d(u,v) > 1$ for all $u,v\in V$. Construct a tree $T$ by:
\begin{enumerate}
\item Choose randomly a permutation $\pi$ of $V$.
\item Choose $\beta \in [0,1]$ uniformly at random.
\item Set $D_\delta = \{V\}$.
\item For $i = \delta -1$ to $0$ do:
\begin{enumerate}
\item Assign every node to the first node according to $\pi$ that has distance at most $2^\beta \cdot 2^{i-1}$.
\item All nodes assigned to the same node that are currently in the same cluster will form a new cluster in $D_i$.
\end{enumerate}
\item Return tree $T$ where $V(T) = \{s \in D_i : \forall i = 0, \dots, \delta\}$ and $E(T) = \{s_is_{i-1}  \in D_{i}\times D_{i-1}: s_{i-1} \subseteq s_i, \forall i = 0,\dots,\delta\}$, and $d(s_is_{i-1}) = 2^i$.
\end{enumerate}
Note that the clusters of $D_0$ are singletons because the bound on the maximum distance is $2^\beta 2^0 \leq 1$. Hence the clusters of $D_0$ can be identified with nodes in $V$. 
\paragraph{Proof of 1.}
Suppose $u,v$ are in the same cluster at level $D_i$ and are separated at level $D_{i-1}$. Let $w$ be the node identified with $u$ and $v$'s cluster $S_w$ at level $D_i$. Then $d(u,v) \leq d(u,w) + d(w,v) \leq 2^\beta 2^{i-1} + 2^\beta 2^{i-1} \leq 2^{i+1}$. On the other hand $d^T(u, S_w) \geq 2^i$ and $d^T(S_w, v) \geq 2^i$ so $d^T(u,v) \geq 2^i + 2^i = 2^{i+1}$. This proves part $1.$ of the theorem.
\paragraph{Proof of 2.}
Observe that if $u,v$ are separated first at level $D_i$ then $d^T(u,v) \leq 2^{i+2}$. Let $S_i$ be the cluster in $D_i$ containing $u,v$. Let $S^u_{i-1}$ be descended cluster of $S_i$ from which $u$ descends and similarly define $S^v_{i-1}$. Then 
\begin{align*}
d^T(u,v) &= d^T(S_i, S^u_{i-1}) + d^T(S_, S^u_{i-1}) + d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v)\\
&= 2^i + 2^i +  d^T(S^u_{i-1}, u) + d^T(S^v_{i-1}, v) \\
&\leq 2^i + 2^i + 2\sum_{j = 1}^{i-1} 2^j \\
&= 2^{i+1} + 2^{i+1} \\
&= 2^{i+2}.
\end{align*}
Now we way that node $c$ {\it settles} $(u,v)$ at level $D_i$ if $c$ is the first node with respect to $\pi$ to which one of $u$ or $v$ gets assigned at level $D_i$. We say that node $c$ {\it cuts} $(u,v)$ at level $D_i$ if exactly one of $u,v$ is at distance at most $2^\beta 2^i$ from $c$. If $u,v$ are separated at level $D_i$ then there exists some $c$ which both settles and cuts $u,v$ at level $D_i$.
\paragraph{}
Note that $$E[d^T(u,v)] \leq \sum_{i=0}^{\delta - 1} \sum_{c \in V} Pr [ \text{$c$ settles (A) and cuts (B) $u,v$ at $D_i$}] 2^{i+2}.$$ The bound simply follows from our observation, and the prior definitions. Recall that $Pr[A \cap B] = P[A\mid B] Pr[B]$.
\paragraph{}
We first analyze $Pr[B]$. Without loss we may assume $d(c,u) \leq d(c,v)$. For $c$ to cut $u,v$ we need $d(c,u) \leq 2^\beta 2^{i-1} < d(c,v)$ by definition. Thus
$$Pr[B] = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{|[2^{i-1},2^i]|} = \frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{2^{i-1}}$$.
Now we need $Pr[A \mid B]$. Let us order the nodes of $V$ according to their distance to $u,v$. That is, $V = \{c_1, \dots, c_n\}$ where $\min\{d(c_i, v), d(c_i, u)\} \leq \min \{d(c_{i+1}, v), d(c_{i+1}, u)\}$ for all $i$. If $c= c_k$ then the probability that $c$ comes before $c_1, \dots, c_{k-1}$ in $\pi$ is at most $\frac{1}{k}$. In order for $c$ (given that $c$ cuts $u,v$) to settle $u,v$ at least that has to happen. Hence
$$Pr[A \mid B] \leq \frac{1}{k}.$$
\paragraph{}
Putting it all together we obtain the following expectation:
\begin{align*}
E[d^T(u,v)] &\leq \sum_{i=0}^{\delta -1} \sum_{k=1}^n Pr [A \mid B]Pr[B] 2^{i+2} \\
&\leq \sum_{k=1}^n \frac{1}{k} \sum_{i=0}^{\delta -1 }  2^{i+2}\frac{| [2^{i-1},2^i] \cap [d(c,u),d(c,v)]|}{2^{i-1}} \\
&= 8 \sum_{k=1}^n \frac{1}{k} \sum_{i=0}^{\delta -1 }  | [2^{i-1},2^i] \cap [d(c,u),d(c,v)]| \\
&\leq 8 d(u,v) \sum_{k=1}^n \frac{1}{k} \\
&= 8 H_n d(u,v) \\
&= O(log n) d(u,v).
\end{align*}
Thus the second claim and therefore the theorem holds.
\end{proof}
\begin{theorem}
There exists a tree $T' = (V,F)$ such that $d^T(u,v) \leq d^{T'}(u,v) \leq 4 d^T(u,v).$
\end{theorem}
\begin{proof}
Pick $v \in V$ with a parent $w \not \in V$ and contract edge $vw$ in $T$. Repeat until all nodes of $T$ are in $V$. Multiply by $4$ the weights on all remaining edges. Call the resulting tree $T'$. Note that clearly $d^{T'}(u,v) \leq 4d^T(u,v)$ as contracting only shortens the distances. Furthermore, in the worst case $d^{T'}(u,v) = 2^i \cdot 4$ if contraction results in contracting $u$ and $v$ right up to the cluster where they were separated (that is, contracting preserves the largest edge distance on the $uv$-path in $T$). But 
$$2^i \cdot 4 = 2^i + 2^i + 2^i + 2^i \geq d(S^u_{i-1}, S_i) + d(S^v_{i-1}, S_i) +  d(S^u_{i-1}, u) + d(S^v_{i-1}, v) = d^T(u,v).$$
Hence the theorem holds.
\end{proof}
\section{Generalized Steiner Network}
\paragraph{}
Here we seek to given an application of the previous theory. In the Generalized Steiner Network problem (GSN), we are given a graph $G=(V,E)$, a distance function $d: E \rightarrow \R_+$, and requirements $r(u,v) \in \N$ for all $u,v \in V$. We are to find an integer capacity function $x : E \rightarrow \N$ of minimum cost $\sum_{e \in E} d(e) x_e$ such that capacities $x$ allow for routing a flow of value $r(u,v)$ from $u$ to $v$.
\paragraph{Proposition 1} Optimizing on $G$ is equivalent to optimizing on $G^M$ for $GSN$ problems. This proposition is immediate from the same arguments as was used for ST.
\paragraph{Proposition 2} GSN is solvable in polynomial time on trees. Since each edge is a cut in the tree, we must set $x_{uv} = r(u,v)$ and this is the best we can do.
\paragraph{Algorithm} Consider the following simple algorithm for GSN:
\begin{enumerate}
\item Find a tree $T$ of $G$ according to previous section's metric embedding algorithms.
\item Solve GSN on $T$ and return the solution.
\end{enumerate}
\begin{theorem}
Algorithm above is a randomized $O(log n)$-approximation algorithm for GSN.
\end{theorem}
\begin{proof}
Let $\bar{x}$ be the solution return by our algorithm. Let $x^*$ be OPT solution. Then $\sum_{uv} d(u,v) \bar{x}_{uv} \leq \sum_{uv} d^T(u,v)\bar{x}_{uv} \leq \sum_{uv} d^T(u,v) x^*_{uv}$. The first inequality follows since $d(u,v) \leq d^T(u,v)$. The second inequality follows since $x^*$ is a feasible solution for $T$. Taking expectation we see:
$$E[\sum_{uv} d(u,v) \bar{x}_{uv}] \leq E[\sum_{uv} d^T(u,v)] \leq O(log n) \sum_{uv} d(u,v) x^*_{uv}$$
with the last inequality following from our theorem.
\end{proof}
\end{document}

>>>>

\lhead{2017-01-19}
\rhead{William Justin Toth CO750-Approximation Algorithms Lecture 5} %CHANGE n to ASSIGNMENT NUMBER ijk TO COURSE CODE
\renewcommand{\headrulewidth}{1pt} %heading underlined
%\renewcommand{\baselinestretch}{1.2} % 1.2 line spacing for legibility (optional)

\begin{document}
\paragraph{}
Before moving on to the next topic, we just want to mention an important theorem that results from the theory built in the previous lectures. It tells us that we can have deterministic approximation algorithms that give just as good bounds as the randomized ones studied previously.
\begin{theorem}
Let $G=(V,E)$ be a graph. Let $d: E \rightarrow \R_+$ be a metric function. Let $\lambda^*_{uv} \geq 0$ be coefficients for all $(u,v) \in V\times V$. There exists a polynomial time algorithm which constructs a weighted tree $T=(V,E)$ such that $d^T(u,v) \geq d(u,v)$ for all $u,v\in V$ and $$\sum_{u,v} \lambda^*_{uv}d^T(u,v) \leq O(\log n) \sum_{u.v} \lambda^*_{uv} d(u,v).$$
\end{theorem}
\begin{proof}
Omitted. May be in a talk at the end of course via some students project. Involves derandomization via the method of conditional expectation.
\end{proof}
\section{Primal - Dual Methods}
\paragraph{}
Primal-Dual methods are a common paradigm in algorithm design. We will begin by showing how they are used to given exact algorithms, and then explain relaxed complementary slackness which allows us to extend the paradigm to give approximation algorithms.
\paragraph{}
Consider the {\it Arborescence Problem} (ARB). In this problem we are given a directed graph $D= (V,E)$, cost function $c: E \rightarrow \R_+$, and a root vertex $r \in V$. We are to find an arborescence $T$ rooted at $r$ which spans $V$ and minimizes $\sum_{e\in E(T)} c_e$.
\paragraph{}
As the name of this method might suggest we will used integrally the linear programming relaxation of this problem:
\begin{align*}
\min \sum_{e \in E} c_e x_e  &\ \\
\text{s.t.} \sum_{e \in \delta^+(S)} x_e &\geq 1 &\forall S \subseteq V\backslash\{r\} \\
x_e &\geq 0 &\forall e \in E,
\end{align*}
and its dual program:
\begin{align*}
\max \sum_{S \subseteq V\backslash\{r\}} y_S &\ \\
\text{s.t.} \sum_{S : e \in \delta^+(S)} y_S &\leq c_e &\forall e \in E \\
y_S &\geq 0 &\forall S \subseteq V\backslash \{r\}.
\end{align*}
Here $\delta^+(S) = \{(u,v) \in E : u \in S, v\not\in S\}$ denotes set of arcs outgoing from $S$.
\paragraph{}
Our primal-dual algorithm for (ARB) operates as follows (start by setting $x=0$, $y=0$, for implementation track only non-negative $y$ to avoid exponential runtime): 
\begin{enumerate}
\item While $x$ is not feasible for the primal:
	\begin{enumerate}
	\item Increase $y_S$ uniformly for all ``strongly connected components" (with respect to $x$) $S$ where $x(\delta^+(S)) = 0$ until some constraints for an edge $e$ becomes tight.
	\item Set $x_e=1$ for tight edge $e$ in previous step.
	\end{enumerate}
\item Deletion Step: Consider each edge $e$ where $x_e = 1$ in reverse order from the order they were set to $1$ and set $x_e = 0$ as long as this does not cause the resulting $x$ to become infeasible.
\item Return $x$.
\end{enumerate}
\begin{theorem}
The previous algorithm is an exact polynomial time algorithm for (ARB).
\end{theorem}
\begin{proof}
When the algorithm terminates we have that if $x_e > 0$ then $x_e = 1$  and so the dual constraint is tight, that is:
$$\sum_{S : e\in\delta^+(S)} y_S = c_e.$
This is immediate from our construction of $x$. Hence the primal complementary slackness conditions hold.
\paragraph{}
For the dual complementary slackness conditions, suppose there exists $S$ such that $y_S > 0$, yet $\sum_{e\in \delta^+(S)} x_e \geq 2$. Let $e, \bar{e} \in \delta^+(S)$ and suppose without loss of generality that $\bar{e}$ was added before $e$ by the algorithm. Since we did not delete $e$, there exists a node $v \in S$ that uses $e$ to reach $r$. But since $S$ is strongly connected, no nodes in $S$ need $\bar{e}$ to reach $r$ (they can go $v$ then eventually through $e$), so we could have deleted $\bar{e}$ in the Deletion Step. A contradiction.
\end{proof}
\end{document}
